mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-10 Draft3.Rmd
custom operators

@@

\pkg{mathml} is a package for pretty mathematical representation of R functions or algorithms in data analysis, scientific reports and interactive web content. The features currently supported are listed below, following the order proposed by [@murrell2000].

@@

## Arithmetic operators
The \pkg{mathml} package provides the graphical representation for addition, subtraction, multiplication and division. For instance:

```{r, results="asis", echo=TRUE}
catmath(x + y - z)
```

In addition, with \pkg{mathml} it is also possible to combine `minus` and `plus` and get a new operator. However, in R\ the operator \(\pm\) is not present and this leads to some interesting questions. If we let R\ present the expression, this is what we get:

```{r, results='asis', echo=TRUE}
# Note wrong precedence of custom operator
catmath(mean(m) %+-% 1.96 * s/sqrt(N))
```

As can be seen, the new custom operator takes precedence over multiplication, as R\ automatically puts `mean(m) %+-% 1.96` in parentheses. Unfortunately, this is not consistent with the rules of mathematics. As the reader well knows, it is not possible in R\ to change the precedence of custom operators. This leaves two possible solutions. The first involves a mathematically unnecessary parenthesis around `(1.96 * s/sqrt(N))`. The second is more elegant and mathematically consistent: here `1.96 * s/sqrt(N)` is put in curly brackets and mathematical formality is respected. In both solutions R\ is able to produce correct output once the order of operations is made explicit.

```{r, results='asis', echo=TRUE}
# This is what we actually want
catmath(mean(m) %+-% (1.96 * s/sqrt(N)))

# Without parentheses
catmath(mean(m) %+-% {1.96 * s/sqrt(N)})
```

>>>>

\pkg{mathml} is an R\ package for pretty mathematical representation of
R\ functions and objects in data analysis, scientific reports and interactive
web content. The features currently supported are listed below, roughly following
the order proposed by [@murrell2000].

## Basic calculations

Basic arithmetic operators and parentheses are translated as they are, as
illustrated below.

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
mathout(term)
```

For multiplications involving only numbers and symbols, the multiplication sign
is omitted. This heuristic does not always produce the desired result;
therefore, \pkg{mathml} defines the extra functions `dot/2`, `nodot/2`,
and `times/2` that produce the respective multiplication signs.

```{r, results="asis"}
term <- quote(dot(d, e) + nodot(f, g + h) + times(X, Y))
mathout(term)
```

Note that for the standard operators with known precedence, \pkg{mathml} is able
to detect if parentheses are needed; for example, a parenthesis is automatically
put around `g + h` in the `nodot`-example and around `n*o` above. However, we
also note an unecessary parenthesis around `l + m` above. This parenthesis
results is a consequence of `quote(a^(b + c))` actually producing a nested
R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

```{r}
term <- quote(a^(b + c))
paste(term)
```

This feature is unfortunate because the extra parenthesis around `b + c` is not
needed and yields to the extra parenthesis. The preferred result is obtained by
using the functional representation `quote('^'(k, l + m))` of the power, or, as
a workaround, curly braces (see last summand above).

## Custom operators

While the parentheses typically follow the precedence in standard infix
operators, undesirable results may be obtained in custom operators.

```{r, results='asis'}
term <- quote(mean(X) %+-% 1.96 * s / sqrt(N))
mathout(term)
```

As can be seen, \pkg{mathml} reminds the user that it is not possible to change
the precedence of custom operators in R, and that expressions with such
operators are evaluated strictly from left to right. As in the
previous example, the solution is to either use the functional form of the
operator or a curly brace to enforce the correct operator precedence.

```{r, results='asis'}
term <- quote(`%+-%`(mean(X), 1.96 * s / sqrt(N)))
term <- quote(mean(X) %+-% {1.96 * s / sqrt(N)})
mathout(term)
```

@@

@@
